Read Me

### 2023-05-05 code/analysis_code/1_EDA.Rmd
submitting final changes

```{r message=FALSE, warning=FALSE}
library(here)
library(spdep)
library(GGally)
library(skimr)
library(ggfortify)
library(MASS)
library(ggcorrplot)
library("FactoMineR")
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
library(ggpubr)
```

@@

```{r}
states <- data.frame(state.name)
states$Region <- case_when(
  states$state.name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania") ~ "Northeast",
  states$state.name %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin") ~ "Midwest",
  states$state.name %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia") ~ "South",
  states$state.name %in% c("Arizona", "California", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming") ~ "West"
)

@@

Let's create a variable for mortality rate ratio of Black/ White
```{r}
race_sub <- race_sub %>%
  group_by(State) %>%
  mutate(MRR = `Death Rate (estimated)`[Race == "Black"] / `Death Rate (estimated)`[Race == "White"])
```

Let's start by looking at the overall HIV-related death rate by race in the United States.
```{r}
table1 <- race_sub %>% 
  group_by(Race) %>% 
  summarize(tot_deaths = sum(`Death Cases`),
            tot_population = sum(`Population`),
            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
  ungroup() %>% 
  mutate(ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
  kable(col.names = c("Race/ Ethnicity",
                      "HIV-Related Deaths (n)",
                      "Population \n (>12 years old)",
                      "Death Rate \n (per 100,000)",
                      "Mortality Rate Ratio"),
        caption = "<center>HIV-Related Deaths by Race/ Ethnicity, United States, 2020</center>",
        align = "c") %>%
  kable_minimal() %>%
  kable_styling(font_size = 30)%>%
gsub("font-size: initial !important;",
"font-size: 30pt !important;",
.)  %>% 
  landscape()

@@

```{r}
region_table <- race_sub %>% 
  group_by(Region, Race) %>% 
  summarize(tot_deaths = sum(`Death Cases`),
            tot_population = sum(`Population`),
            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
  ungroup() %>% 
  group_by(Region) %>% 
  mutate(ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
  kable(col.names = c("Region",
                      "Race/ Ethnicity",
                      "HIV-Related Deaths (n)",
                      "Population \n (>12 years old)",
                      "Death Rate \n (per 100,000)",
                      "Mortality Rate Ratio"),
        caption = "<center>HIV-Related Deaths by Region, United States, 2020</center>") %>% 
  kable_minimal() %>%
  kable_styling(font_size = 30) %>%
gsub("font-size: initial !important;",
"font-size: 30pt !important;",.)  %>% 
  landscape()

@@

Next, let's take a look at our response variable
Univariate Analysis

Let's take a look at some box plots of death rate by races and state. 
We'll start by looking at death rate given by the AIDSVu dataset. 
```{r}
race_sub %>% 
  ggplot() +
  geom_boxplot(aes(y = `Death Rate (estimated)`, x = Race, group = Race)) + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio, United States, 2020", y = "Ratio", x = "Race/ Ethnicity") + theme(legend.position = "none") 
```

Histogram
```{r}
race_sub %>% 
  filter(Race == "Black") %>% 
  ggplot() +
  geom_histogram(aes(x = `MRR`), bins = 10) + 
  theme(legend.position = "none") +
  labs(title = "Histogram of Black-to-White HIV-Related Mortality Rate Ratio, United States, 2020", y = "n", x = "Ratio") + theme_classic()
ggsave(here("results/MRR_hist.png"), units="in", width=10, height=7, dpi=300)

@@

race_sub %>% 
  ggplot() +
  geom_boxplot(aes(y = `Death Rate (estimated)`, x = Race, group = Race)) + 
  theme(legend.position = "none") +
  labs(title = "HIV-Related Death Rate, by Race, United States, 2020", y = "Rate (per 100,000)", x = "Race") + theme_classic()
ggsave(here("results/dr_race_bp.png"), units="in", width=10, height=5, dpi=300)

race_sub %>% 
  ggplot() +
  geom_bar(aes(y = `Death Rate (estimated)`, x = `State Abbreviation`, group = `State Abbreviation`, fill = Race),
           stat = "identity", position = "dodge") + theme_classic() + 
  theme(legend.position = "bottom") +
  labs(title = "HIV-Related Death Rate of Black and White Persons, by State, United States, 2020", y = "Rate (per 100,000)", x = "State")
ggsave(here("results/dr_state.png"), units="in", width=12, height=5, dpi=300)

race_sub %>% 
  ggplot() +
  geom_boxplot(aes(y = `Death Rate (estimated)`, x = `Region`, group = `Region`)) + theme_classic() + 
  theme(legend.position = "bottom") +
  labs(title = "HIV-Related Death Rate of Black and White Persons, by Region, United States, 2020", y = "Rate (per 100,000)", x = "Region") + facet_wrap(.~ Race)
ggsave(here("results/dr_region.png"), units="in", width=12, height=5, dpi=300)
```

Let's move on to
Bivariate Analysis 
We will deem variables with a correlation coeffiencient of |0.15| as important, and will include these in PCA
```{r}
bi_1 <- ggplot(race_sub, aes(y = MRR, x = `Percent Living in Poverty`)) +
  geom_smooth(method = "lm") + 
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR)
  vs. Percent Living in Poverty") + 
  stat_cor(method = "pearson") 
#  pass 
#ggsave(here("results/bi_pov.png"), units="in", width=10, height=7, dpi=300)
```

@@

```{r}
pos_predictors <- race_sub %>%
  ungroup() %>% 
  select(c("Median Household Income",
           "Percent Living in Poverty",
           "Percent Without an Internet subscription",
           "Percent Uninsured Under 65 Years Old",
           "Percent Living with Food Insecurity",
           "Percentage Diagnosed Late")) %>% 
  names()
```

>>>>

```{r message=FALSE, warning=FALSE}
library(here)
library(ggfortify)
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
library(ggpubr)
```

@@

Creating variable for region
```{r}
states <- data.frame(state.name)
states$Region <- case_when(
  states$state.name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania") ~ "Northeast",
  states$state.name %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin") ~ "Midwest",
  states$state.name %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia") ~ "South",
  states$state.name %in% c("Arizona", "California", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming") ~ "West"
)

@@

Let's start by looking at the overall HIV-related death rate by race in the United States.
```{r}
table1 <- race_sub %>% 
  group_by(Race) %>% 
  summarize(tot_deaths = sum(`Death Cases`),
            tot_population = sum(`Population`),
            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
  ungroup() %>% 
  mutate(ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
  kable(col.names = c("Race/ Ethnicity",
                      "HIV-Related Deaths (n)",
                      "Population \n (>12 years old)",
                      "Death Rate \n (per 100,000)",
                      "Mortality Rate Ratio"),
        caption = "<center>HIV-Related Deaths by Race/ Ethnicity, United States, 2020</center>",
        align = "c", format.args = list(big.mark = ",")) %>%
  kable_minimal() %>%
  kable_styling(font_size = 50)%>%
gsub("font-size: initial !important;",
"font-size: 50pt !important;",
.)  %>% 
  landscape()

@@

```{r}
region_table <- race_sub %>% 
  group_by(Region, Race) %>% 
  summarize(tot_deaths = sum(`Death Cases`),
            tot_population = sum(`Population`),
            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
  ungroup() %>% 
  group_by(Region) %>% 
  mutate(ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
  kable(col.names = c("Region",
                      "Race/ Ethnicity",
                      "HIV-Related Deaths (n)",
                      "Population \n (>12 years old)",
                      "Death Rate \n (per 100,000)",
                      "Mortality Rate Ratio"),
        caption = "<center>HIV-Related Deaths by Region, United States, 2020</center>", format.args = list(big.mark = ",")) %>% 
  kable_minimal() %>%
  kable_styling(font_size = 50) %>%
gsub("font-size: initial !important;",
"font-size: 50pt !important;",.)  %>% 
  landscape()

@@

Next we'll look into the distribution of our response variable
```{r}
race_sub %>% 
  filter(Race == "Black") %>% 
  ggplot() +
  geom_histogram(aes(x = `MRR`), bins = 10) + 
  theme(legend.position = "none") +
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio, United States, 2020", y = "n", x = "Ratio") + theme_classic()
ggsave(here("results/MRR_hist.png"), units="in", width=10, height=5, dpi=300)

@@

```

Let's also look at the distribution of HIV-related death rate
```{r}
race_sub %>% 
  ggplot() +
  geom_histogram(aes(x = `Death Rate (estimated)`, fill = Race), bins = 20) + 
  theme(legend.position = "none") +
  labs(title = "HIV-Related Death Rate, United States, 2020", y = "n", x = "Deaths per 100,000") +
  theme_classic()
ggsave(here("results/DR_hist.png"), units="in", width=10, height=5, dpi=300)

```

Let's take a look at MRR by state and region
```{r}
race_sub %>% 
  ggplot() +
  geom_bar(aes(y = `MRR`, x = `State Abbreviation`, group = `State Abbreviation`),
           stat = "identity", position = "dodge") +
  theme_classic() + 
  theme(legend.position = "bottom") +
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio, by State, United States, 2020", y = "Ratio", x = "State")
ggsave(here("results/mrr_state.png"), units="in", width=12, height=5, dpi=300)

race_sub %>% 
  ggplot() +
  geom_boxplot(aes(y = `MRR`, x = `Region`, group = `Region`)) +
  theme_classic() + 
  theme(legend.position = "bottom") +
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio, by Region, United States, 2020", y = "Ratio", x = "Region")
ggsave(here("results/mrr_region.png"), units="in", width=12, height=5, dpi=300)
```

Let's move on to
Bivariate Analysis 
We will deem variables with a correlation coefficient of |0.15| as important, and will include these in PCA
```{r}
bi_1 <- ggplot(race_sub, aes(y = MRR, x = `Percent Living in Poverty`)) +
  geom_smooth(method = "lm") + 
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR)
  vs. Percent Living in Poverty") + 
  stat_cor(method = "pearson") 
#  pass 
#ggsave(here("results/bi_pov.png"), units="in", width=10, height=7, dpi=300)
```

@@

```{r}
pos_predictors <- race_sub %>%
  dplyr::select(c("Median Household Income",
           "Percent Living in Poverty",
           "Percent Without an Internet subscription",
           "Percent Uninsured Under 65 Years Old",
           "Percent Living with Food Insecurity",
           "Percentage Diagnosed Late")) %>% 
  names()
```

### 2023-05-05 code/processing_code/1_cleaning_2020.Rmd
submitting final changes

deaths_sex <- deaths %>%
  select(1:4, contains("male"), -contains("and"), -ends_with("Stability"))

deaths_race <-   deaths %>%
  select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")), -ends_with("Stability"))

@@

Next, let's clean the dataframe for New diagnoses. We'll select for variables of interest and subset.
```{r}
new_dx <- new_dx %>% 
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  select(1:3, contains(c("male", "Black", "White", "Hispanic", "Multiple")), -contains(c("and", "Stability")))

sex_dx <- new_dx %>% 
  select(1:7)

new_dx <- new_dx %>% 
  select(-contains(c("male", "Year")))
```

@@

I'll select only the columns of interest, and rename some of them so they are a bit easier to work with.
```{r}
race_pop <- race_pop %>% 
  select(c(Location, starts_with(c("Total:", "Black or African American alone", "Hispanic", "White alone", "Population of two or more races:")))) %>% 
  rename(c(Hispanic = `Hispanic or Latino`,
         White = `White alone`,
         Black = `Black or African American alone`,
         Multiple = `Population of two or more races:`))
```

@@

total_pop <- race_pop %>%
  select(1:2)
```

@@

Cleaning the file `Received Care`. I do some cleaning and merge appropriate dataframes, and use some base R to add a couple columns to `sex_df`.
```{r}
rec_care <- rec_care %>%
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  select(1:3, contains("Percent of")) %>% 
  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))

@@

rec_care <- rec_care %>%
  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple")))

race_df <- inner_join(race_df, rec_care)
```
Repeat for file on `Viral Suppression`
```{r}
v_sup <- v_sup %>%
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  select(1:3, contains("Percent of")) %>% 
  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))

@@

v_sup <- v_sup %>%
  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple")))
race_df <- inner_join(race_df, v_sup)
```

@@

prev_sex <- prev %>%
  select(1:4, contains("male"), -contains("and"), -ends_with("Stability"))

@@

prev_race <-   prev %>%
  select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")), -contains("and"), -ends_with("Stability"))

@@

```{r}
no_internet <- internet %>% 
  select(c(1, "Without an Internet subscription")) %>% 
  slice(c(-1:-8, -48:-52,-263:-267)) %>% 
  filter(!grepl(",", `Without an Internet subscription`)) %>% 
  rename(`Percent Without an Internet subscription` = `Without an Internet subscription`)
```

@@

no_internet <- no_internet %>% 
  select(State, `Percent Without an Internet subscription`) %>% 
  drop_na()

@@

```{r}
race_df <- full_join(race_df, no_internet)
sex_df <- full_join(sex_df, no_internet)
```

We have some more cleaning to, but let's save these files for now. 
```{r}
saveRDS(sex_df, here("data/processed_data/sex_df.rds"))
saveRDS(race_df, here("data/processed_data/race_df.rds"))
```

>>>>

deaths_sex <- deaths %>%
  dplyr::select(1:4, contains("male"), -contains("and"), -ends_with("Stability"))

deaths_race <- deaths %>%
  dplyr::select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")), -ends_with("Stability"))

@@

Next, let's clean the dataframe for New diagnoses. We'll select for variables of interest and subset.
```{r}
new_dx <- new_dx %>% 
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  dplyr::select(1:3,
                contains(c("male", "Black", "White", "Hispanic", "Multiple")),
                -contains(c("and", "Stability")))

sex_dx <- new_dx %>% 
  dplyr::select(1:7)

new_dx <- new_dx %>% 
  dplyr::select(-contains(c("male", "Year")))
```

@@

I'll select only the columns of interest, and rename some of them so they are a bit easier to work with.
```{r}
race_pop <- race_pop %>% 
  dplyr::select(c(Location, starts_with(c("Total:", "Black or African American alone", "Hispanic", "White alone", "Population of two or more races:")))) %>% 
  rename(c(Hispanic = `Hispanic or Latino`,
         White = `White alone`,
         Black = `Black or African American alone`,
         Multiple = `Population of two or more races:`))
```

@@

total_pop <- race_pop %>%
  dplyr::select(1:2)
```

@@

Cleaning the file `Received Care`. I do some cleaning and merge appropriate dataframes, and use some base R to add a couple columns to `sex_df`.
```{r}
rec_care <- rec_care %>%
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  dplyr::select(1:3, contains("Percent of")) %>% 
  dplyr::select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))

@@

rec_care <- rec_care %>%
  dplyr::select(1:3, contains(c("Hispanic", "Black", "White", "Multiple")))

race_df <- inner_join(race_df, rec_care)
```
Repeat for file on `Viral Suppression`
```{r}
v_sup <- v_sup %>%
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  dplyr::select(1:3, contains("Percent of")) %>% 
  dplyr::select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))

@@

v_sup <- v_sup %>%
  dplyr::select(1:3, contains(c("Hispanic", "Black", "White", "Multiple")))
race_df <- inner_join(race_df, v_sup)
```

@@

prev_sex <- prev %>%
  dplyr::select(1:4, contains("male"),
                -contains("and"),
                -ends_with("Stability"))

@@

prev_race <-   prev %>%
  dplyr::select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")),
                -contains("and"),
                -ends_with("Stability"))

@@

```{r}
no_internet <- internet %>% 
  dplyr::select(c(1, "Without an Internet subscription")) %>% 
  slice(c(-1:-8, -48:-52,-263:-267)) %>% 
  filter(!grepl(",", `Without an Internet subscription`)) %>% 
  rename(`Percent Without an Internet subscription` = `Without an Internet subscription`)
```

@@

no_internet <- no_internet %>% 
  dplyr::select(State, `Percent Without an Internet subscription`) %>% 
  drop_na()

@@

```{r}
race_df <- full_join(race_df, no_internet)
sex_df <- full_join(sex_df, no_internet)
```
We have some more cleaning to, but let's save these files for now. 
```{r}
saveRDS(sex_df, here("data/processed_data/sex_df.rds"))
saveRDS(race_df, here("data/processed_data/race_df.rds"))
```
